\blue{Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media - change the title to something close to the main topic of the manuscript ... something with viscous flow instability or viscous fingering)}

### 2017-05-07 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Updates: starting revising Introduction

Due to complex and/or heterogeneous geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial to understand the aforementioned processes. Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and the flow regimes in capillary media. A comprehensive review of mechanisms of fingering (i.e., fluid instability) and forces acting on the interface between two immiscible fluids can be found in \citet{homsy_1987}. \citet{adler_1988} described both microscopic and macroscopic scale darcy-equations \red{(BE MORE SPECIFIC ON THEIR CONTRIBUTION TO THE STATE-OF-THE-ART OF THE FIELD!!)}. Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand, macroscopic instabilities involve two-phase flow system \citep{bottoni_1992}. This paper will focus on microscopic instabilities and phenomena in two-phase systems. A local flow instability that is the focus of this work is due to viscous and stress forces and often named as viscous instabilities or \textit{viscous fingering} \red{(TRY TO MERGE THESE 2 SENTENCES!)}. 
In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these two fluids moves along the geological formation, creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and, may lead to early breakthrough of water into neighbour production wells. Viscous instabilities are mainly governed by the ratio of mobility (MR) between the displacing and the displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (i.e., large spatial permeability gradient) , gravitational forces, anisotropic dispersion, velocity dependence of dispersion, non-monotonic viscosity profiles, etc.

\medskip 
Conservation of mass along with the Darcy's law are the governing equations that describe the flow dynamics in porous media. Several computational methods have been developed to accurately represent force balances in Darcean flows.
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). 
%The FDM discretization is based upon the differential form of the PDE to be solved. 
%Analytic models can be used to analyze reservoir performance as long as the reservoir is not highly heterogeneous. However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geological domains. FEM discretization is based upon a piecewise representation of the solution space  and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times.

\medskip
Among FEM-based formulations for porous media, the control volume finite element methods, ~\citep[CVFEM,][]{fung_1992} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In the CVFEM pressure is interpolated using piecewise linear finite element basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$) 
while saturation is represented in node centered CV basis functions respectively, and the saturation equations are solved explicitly after solving for pressure at each non-linear iteration.

>>>>

Due to complex and/or heterogeneous geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial to understand the aforementioned processes. Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and the flow regimes in capillary media. A comprehensive review of mechanisms of fingering (i.e., fluid instability) and forces acting on the interface between two immiscible fluids can be found in \citet{homsy_1987}. \citet{adler_1988} described both microscopic and macroscopic scale darcy-equations \red{(BE MORE SPECIFIC ON THEIR CONTRIBUTION TO THE STATE-OF-THE-ART OF THE FIELD!!)}. Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand, macroscopic instabilities involve two-phase flow system \citep{bottoni_1992}. This paper will focus on microscopic instabilities and phenomena in two-phase systems. A local flow instability that is the focus of this work is due to viscous and stress forces and often named as viscous instabilities or \textit{viscous fingering} \red{(TRY TO MERGE THESE 2 SENTENCES!)}.

In oil and gas reservoir exploration, viscous and density instabilities are relatively common during water-injection processes. As water and oil interacts, the interface between these two fluids moves along the geological formation, creating an uneven or fingered flow profile (see Fig.~\ref{fig:simple_case}). Viscous fingering results in inefficient flow sweeping which can bypass significant quantities of recoverable oil and, may lead to early breakthrough of water into neighbour production wells. Viscous instabilities are mainly governed by the ratio of mobility (MR) between the displacing and the displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (i.e., large spatial permeability gradient) , gravitational forces, anisotropic dispersion, velocity dependence of dispersion, non-monotonic viscosity profiles, etc.

\medskip 
Conservation of mass and Darcy's law are the governing equations that describe the flow dynamics in porous media. Several computational methods have been developed to accurately represent force balances in Darcean flows. Finite difference methods (FDM) have been extensively used in most industry-standard reservoir simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications. However, FDM are often limited to relatively simple geometries representing idealised geological formations through structured quadilateral (2-D) and hexahedral (3-D) grid cells of low quality~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}.

The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient than FDM schemes when they are used to solve complex geometries. FEM discretisation is based upon piecewise representation of the solution space and the computational domain is divided into smaller subdomains (elements) where the solution in each element is constructed from basis functions. %Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times. %At this point let us remind that a comparison of computaional methods is difficult and depends always on the test-case.
%The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to be more efficient compared with FDM schemes when they are used to solve complex geometrical domains. FEM discretization is based upon a piecewise representation of the solution in terms of specified basis functions. The computational domain is divided into smaller domains (elements) and the solution in each element is constructed from the basis functions. Advantages of FEM over FVM are the better description of the geometry and complex shapes, better treatment of fluid flow and more accurate field-solution calculations using a smaller number of nodes, less memory space and shorter run times.

Among FEM-based formulations for porous media, the control volume finite element methods, ~\citep[CVFEM,][]{fung_1992} has been largely used as it can guarantee local mass conservation and high-order numerical accuracy as well as being able to use tetrahedral geometry-conforming elements. In standard CVFEM formulations, pressure and velocity are interpolated using piecewise linear finite element basis functions, %$\underline{\bf v}$ and $\underline{\bf p}$) 
while saturation is represented in node centered CV basis functions.  Saturation equations are solved explicitly after solving for pressure at each non-linear iteration~\citep[a detailed description of the implicit pressure explicit saturation, IMPES, formulation can be found in][]{lux_2012,chen_2006}.
